Peer Assessment 1: Activity monitoring data
===========================================

###Introduction


It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the Уquantified selfФ movement Ц a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

###Data

The data for this assignment can be downloaded from the course web site: 

- Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)  

The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
- date: The date on which the measurement was taken in YYYY-MM-DD format

- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

###Loading and preprocessing the data

```{r}
setwd("~/R/repdata-data-activity")
data<-read.csv("activity.csv", stringsAsFactors=FALSE)
data$date<-as.Date(data$date)
str(data)
```

###What is mean total number of steps taken per day?

```{r}
library(reshape2)
meltdata<-melt(data,id=c("date","interval"),measure.vars="steps")
dailydata<-dcast(meltdata, date~variable,sum)
hist(dailydata$steps, breaks=15,col="cyan", 
     main="Histogram of total number of steps taken per day")
abline(v=median(dailydata$steps,na.rm=TRUE),col='red',lwd=4)

```

**Mean and median of total number of steps taken each day:**

```{r}
median(dailydata$steps,na.rm=TRUE)
mean(dailydata$steps,na.rm=TRUE)
```

###What is the average daily activity pattern?

```{r}
intdata<-dcast(meltdata,interval~variable, mean, na.rm=TRUE)
plot(intdata, type="l", lwd=1, main="Average daily activity pattern")
```

**Interval with the maxmum average number of steps:**

```{r}
intdata[intdata$steps==max(intdata$steps),]
```

###Imputing missing values

Total number of missing values:
```{r}
sum(is.na(data))
```

Imputing missing values with the average number of steps for particular 5-minute interval:

```{r}
impdata<-data
for (i in 1:nrow(data))
  {
    if (is.na(data$steps[i]))
    impdata$steps[i]<-intdata[intdata$interval==impdata$interval[i],2]
  }
```

Checking if any mising value left:

```{r}
sum(is.na(impdata))
```

Analysing total number of steps taken each day for imputed dataset:

```{r}
meltdata1<-melt(impdata,id=c("date","interval"),measure.vars="steps")
dailydata1<-dcast(meltdata1, date~variable,sum)
hist(dailydata1$steps, breaks=15,col="cyan", 
     main="Histogram of total number of steps taken per day (after imputing NAs)")
abline(v=median(dailydata1$steps,na.rm=TRUE),col='red',lwd=4)
median(dailydata1$steps,na.rm=TRUE)
mean(dailydata1$steps,na.rm=TRUE)
```

Comparing original and imputed datasets:

```{r, fig.width=10}
par(mfrow=c(1,2))
hist(dailydata$steps, breaks=15,col="cyan", 
     main="Total number of steps per day")
abline(v=median(dailydata$steps,na.rm=TRUE),col='red',lwd=4)
hist(dailydata1$steps, breaks=15,col="cyan", 
     main="Total number of steps per day (imputed dataset)")
abline(v=median(dailydata1$steps,na.rm=TRUE),col='red',lwd=4)
```

  The historgams above demonstrate that imputing missing values in the dataset didn't change mean and median significantly, but they became closer to each other. The histigorgam changed a little, as in the second dataset more days have total number of steps close to average. Such effect is natural, considering the strategy used for imputing missing values.

###Are there differences in activity patterns between weekdays and weekends?

```{r, fig.width=10}
library(lattice)
impdata$weekday<-factor(ifelse(weekdays(impdata$date) %in% c("понедельник","вторник", "среда", "четверг", "п€тница"),"weekday","weekend"))
meltdata2<-melt(impdata,id=c("date","interval","weekday"),measure.vars="steps")
intdata2<-dcast(meltdata2, weekday+interval~variable,mean)
xyplot(steps~interval | weekday, data=intdata2, type="l", layout=c(1,2))
```

From the figures above we can see that indeed activity patterns are a bit diffrent for weekdays and weekends: on weekend the given individual takes less steps in the morning, but more steps during the day. Interesting, that maximum of steps taken still takes place in the morning, regardless of the weekday.

